Warming Degrees and Flowering Trees

### 2025-04-04 paper/paper.qmd
committing prediction and diagnostic models to qmd.qmd
added more model visualizations

>>>>

```{python}
#| label: fig-train-plt
#| fig-cap: "Trained Predictions Visualization"
# Test OLS model (nonrobust)
model_test = model_0

# Prepare test data
X_test = sm.add_constant(test_df['march_temp_c'])
y_test = test_df['flower_doy']

# Predict on test data
y_pred = model_test.predict(X_test)

# Evaluation metrics
rmse = np.sqrt(mean_squared_error(y_test, y_pred))
mae = mean_absolute_error(y_test, y_pred)
r2 = r2_score(y_test, y_pred)

# Plot actual vs predicted
plt.figure(figsize=(10, 6))
plt.plot(test_df['year'], y_test, label='Actual Blooming DoY', marker='o')
plt.plot(test_df['year'], y_pred, label='Predicted Blooming DoY', marker='x')
plt.xlabel("Year")
plt.ylabel("Blooming Day of Year")
plt.title("Actual vs Predicted Sakura Blooming (2001–2019)")
plt.legend()
plt.grid(True)
plt.show()

# Evaluation metrics print
print("Evaluation on Test Set (2001–2019):")
print(f"RMSE: {rmse:.2f}")
print(f"MAE:  {mae:.2f}")
print(f"R²:   {r2:.3f}")
```

```{python}
#| label: fig-diag-results
#| fig-cap: "Diagnostic Results for Cherry Blossom Flowering Day vs. March Temperature"

# Create diagnostic plots for the OLS model
fig = plt.figure(figsize=(20, 15))

# Residual vs Fitted Plot
ax1 = fig.add_subplot(2, 2, 1)
residuals = model_0.resid
fitted = model_0.fittedvalues
sns.residplot(x=fitted, y=residuals, lowess=True, scatter_kws={'alpha': 0.5}, line_kws={'color': 'red', 'lw': 1}, ax=ax1)
ax1.set_title('Residuals vs Fitted')
ax1.set_xlabel('Fitted values')
ax1.set_ylabel('Residuals')
ax1.axhline(y=0, color='k', linestyle='-', alpha=0.3)

# QQ Plot for Normality
ax2 = fig.add_subplot(2, 2, 2)
qqplot(model_0.resid, line='s', ax=ax2)
ax2.set_title('Normal Q-Q')
ax2.set_xlabel('Theoretical Quantiles')
ax2.set_ylabel('Standardized Residuals')

# Scale-Location Plot (Square root of absolute residuals vs fitted)
ax3 = fig.add_subplot(2, 2, 3)
sqrt_abs_resid = np.sqrt(np.abs(residuals))
sns.regplot(x=fitted, y=sqrt_abs_resid, lowess=True, scatter_kws={'alpha': 0.5}, line_kws={'color': 'red', 'lw': 1}, ax=ax3)
ax3.set_title('Scale-Location')
ax3.set_xlabel('Fitted values')
ax3.set_ylabel('$\\sqrt{|Standardized Residuals|}$')

plt.tight_layout()

# Leverage vs. Standardized Residuals
ax4 = fig.add_subplot(2, 2, 4)
infl = model_0.get_influence()
hatvals = infl.hat_matrix_diag
stud_resid = infl.resid_studentized_external

# Determine leverage threshold
p = len(model_0.params)  # number of parameters
leverage_threshold = 2 * p / len(hatvals)

# Plot points
ax4.scatter(hatvals, stud_resid, alpha=0.6)
ax4.axhline(y=0, color='k', linestyle='-', alpha=0.3)

# Add reference lines
ax4.axhline(y=2, color='g', linestyle='--', label='Std. Residuals = 2')
ax4.axhline(y=-2, color='g', linestyle='--')
ax4.axvline(x=leverage_threshold, color='b', linestyle='--', 
           label=f'Leverage threshold ({leverage_threshold:.3f})')

# Highlight and label influential points (leverage > threshold or std. residuals > 2)
influential_points = []
for i in range(len(hatvals)):
    if (abs(stud_resid[i]) > 2) or (hatvals[i] > leverage_threshold):
        ax4.text(hatvals[i], stud_resid[i], 
                f" {i} ({hatvals[i]:.3f}, {stud_resid[i]:.2f})", 
                fontsize=9, color="red")
        influential_points.append(i)

ax4.set_title('Leverage vs. Standardized Residuals')
ax4.set_xlabel('Leverage (hat values)')
ax4.set_ylabel('Studentized Residuals')
ax4.legend(loc='upper right', fontsize=8)

plt.tight_layout()
plt.show()
```
